MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2021-09-29 main/MSstatsPTM_Main.tex
Made updates to both main and supp

\title{Statistical methods for relative quantification of post-translational modifications in global proteomics experiments}

@@

The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs). Experiments targeting PTMs face several specific challenges. These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression. Due to these challenges, a robust approach to estimate relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment. We propose a general statistical model and workflow that is both reproducible and comprehensive. The model measures modified and unmodified peptide abundance by summarizing intensities through Tukey’s median polish method. Then a model based on the family of linear mixed-effects models is fit. This model is automatically adjusted to the specific experimental design. Finally, the PTM abundances are adjusted to remove variance from changes in the overall protein. We implement this model in the free and open-source R package MSstatsPTM.

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with limma were also proposed.\cite{Zhu} While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).\cite{Breitwieser:2013} Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.

We propose a general statistical approach, which explicitly characterizes the variations and confounding factors present in bottom-up PTM experiments. The proposed approach is aimed at the detection of quantitative changes in PTMs between conditions utilizing procedures developed for summarization of LC-MS data, quantitative characterization of site-specific PTMs, and adjustment with respect to protein abundance. Quantitative analyses of PTMs often involve comparisons between multiple inter-related conditions of the same biological system. The general statistical framework underlying the proposed approach allows for analyzing experiments with complex designs and different acquisition methods, including label-free, DDA, DIA, and TMT.

The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations. The approach was compared against the commonly applied t-test and limma methods. The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate. The proposed approach is implemented as an open source R package MSstatsPTM, which employs similar input format as in MSstats and MSstatsTMT. \cite{Choi:2014} \cite{Huang:2020}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proposed Approach}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Statistical modeling and parameter estimation}

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. In addition mixture may be included for features acquired via tandem mass tag (TMT) methods. The model parameters are estimated using the split-plot approach as in MSstats, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Detection of changes in PTMs}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Missing Value imputation}

MSstatsPTM gives the ability to impute missing feature intensities if desired. When values are imputed, it is assumed they are missing for reasons of low abundance. Missing features are imputed in each MS run using the Accelerated Time Failure model.\cite{Tukey:1977} In order to impute a feature's missing values, the feature must be present in at least one MS Run. If the feature is not present across all MS Runs, the value will be left missing. Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers. Missing value imputation is done separately for both the PTM and global protein datasets.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Extension to TMT experiments}

The statistical modeling approaches discussed above can also be extended to Tandem Mass Tag (TMT) labeling methods. TMT experiments introduce an additional source of variation in the form of different mixtures. To account for mixture variance, first feature intensities are summarized again using Tukey's Median Polish and missing values are imputed, if desired, then a new linear mixed effects model with an added term for the mixture is included. The workflow and statisical model described generally follow the methods used for modeling proteins in MSstatsTMT. \cite{Huang:2020} These methods are repeated to quantify both modified and unmodified peptide abundance. Once both modified and unmodified peptides are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods described previously.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

\subsection*{Computer simulations}
\label{sec:comp_sim}

The proposed statistical approach methods were evaluated using computer simulation. Specifically, their properties under adjustment with respect to protein abundance were evaluated. Two simulations were ran, one including a high number of features and no missing values, and one with few modified features and including missing values.

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test or limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverage measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Further details of how the simulations were generated and their results can be found in Supplementary Sec. \ref{supp-sec:sim}.

\paragraph*{Computer simulation 1}

In this simulation, a clean dataset was generated with a high number of modified and unmodified peptide features and no missing values. Not accounting for changes in unmodified peptide level resulted in a high false discovery rate for all methods, seen in Figure \ref{fig:fdr_all_models}. When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR. Recall and accuracy of the proposed approach and limma performed similarly, while t-test lagged behind. Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach and limma leveraged all available information, which resulted in improved power with increased number of conditions. Further results are provided in Supplementary Figures \ref{supp-fig:sim1_fpr} to \ref{supp-fig:sim1_recall}.

\paragraph*{Computer simulation 2}

In this simulation real world experimental conditions were introduced, including missing values and low number of modified features. In this case, changes in unmodified peptide level still needed to be accounted for to control the FDR in all methods. Once controlling for changes in unmodified peptide, it is clear that the proposed method performed stronger across all methods when compared to limma and t-test, as seen in Figure \ref{fig:sim_accuracy}. The proposed method well calibrated model accuracy and outperformed the other methods. Additionally, in the recall rate the proposed method showed the strongest performance, even when the number of biological replicates was low. Finally, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, while the other methods showed much wider distributions. Further results are provided in Supplementary Figures \ref{supp-fig:sim2_recall} to \ref{supp-fig:fc_boxplot}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Spike-in experiment}

We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark. Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions. Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change. Two sets of data were acquired for each mixture: modified peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate. These datasets were used as the PTM and global protein data.

@@

In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using limma with protein adjustment discussed in Section~\ref{sec:comp_sim}. Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied. However, using limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue. In this case, using limma would have resulted in missing the majority of differential modified peptides.  In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer afterward.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Re-analysis of published datasets}

\subsubsection*{IpaH7.8}

@@

\subsubsection*{Shigella flexneri}

To evaluate our approach we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri). \cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours). Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications. Two mixtures using 11-plex were ran over the six conditions. The six conditions were split between 11 channels leading to the experimental design being unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.

@@

\subsubsection*{USP30}

This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides. A corresponding mixed effects model was fit per modification and global protein as described previously.

In contrast to other experiments analyzed in this paper, there was no unmodified global protein profiling run performed in this experiment. Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a significant lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified protein and could be adjusted. Not having a separate global profiling run resulted in very low feature counts for the unmodified protein model.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion and discussion}

@@

Overall, the proposed approach balances accuracy and practicality, and enables the analysis of complex experiments in high throughput. Future work is to carry out the inference and testing for not only the relative change of PTM abundance, but also the fraction of the protein that is modified at the particular site (site occupancy, or stoichiometry). We are also interested in characterizing the interplay of PTMs at multiple sites. The proposed statistical methods are implemented as an R package MSstatsPTM available on Bioconductor and Github.

>>>>

\title{MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments}

@@

The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs). Experiments targeting PTMs face several specific challenges. These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression. Due to these challenges, a robust approach to estimate relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment. We propose a general statistical model and workflow that is both reproducible and comprehensive. The model measures modified and unmodified peptide abundance by summarizing intensities through Tukey’s median polish method. Then a model based on the family of linear mixed-effects models is fit. This model is automatically adjusted to the specific experimental design. Finally, the PTM abundances are adjusted to remove variance from changes in the overall protein. We implement this model in the free and open-source R package $MSstatsPTM$.

@@

We propose a general statistical approach, which explicitly characterizes the variations and confounding factors present in bottom-up PTM experiments. The proposed approach is aimed at the detection of quantitative changes in PTMs between conditions utilizing procedures developed for summarization of LC-MS data, quantitative characterization of site-specific PTMs, and adjustment with respect to protein abundance. Quantitative analyses of PTMs often involve comparisons between multiple inter-related conditions of the same biological system. The general statistical framework underlying the proposed approach allows for analyzing experiments with complex designs and different acquisition methods.

The proposed approach was evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations. The approach was compared against the commonly applied t-test and limma methods. The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate. Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT. \cite{Choi:2014} \cite{Huang:2020}

\section{Experimental Procedure}

This section describes the datasets and experiments used to evaluate the proposed method. Further details can be seen in Supplemental Sections \ref{supp-sec:sim} and \ref{supp-sec:bio_investigations}

\subsection*{Computer simulations}
\label{sec:comp_sim_procedure}

The proposed statistical approach methods were evaluated using two computer simulations, with known ground truth. Specifically, their properties under adjustment with respect to protein abundance were evaluated. Two simulations were ran, one including a high number of features and no missing values, and one with few modified features and including missing values.

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test or limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach and limma leverage measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Further details of how the simulations were generated can be found in Supplementary Sec. \ref{supp-sec:sim}.

\paragraph*{Computer simulation 1 - label free}

In this simulation, a clean dataset was generated with a high number of modified and unmodified peptide features and no missing values. Multiple simulations were generated with these parameters, with different numbers of replicates and conditions. 1000 modified peptides were generated, half of which were true positives and the other half true negatives. Of true positives simulated, half of them had their abundance's masked by changes in global protein. Modified peptides and unmodified proteins were both simulated with 10 features.

\paragraph*{Computer simulation 2 - missing values and low replicates}

In this simulation real world experimental conditions were introduced, including missing values and low number of modified features. Missing values and feature count values for this simulation were determined by looking at the averages from the biological experiments in this paper. Modified peptides were simulated with 2 features, while unmodified proteins were still simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were masked completely at random. Adding missing values and few representative modified features provides a more realistic expectation of model performance.

\subsection*{SpikeIn benchmark - KGG enriched}

We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark. Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions. Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. Four mixes of spike in peptides and Human Lysate were created to create conditions with known fold change. Two sets of data were acquired for each mixture: modified peptide data, including the spike in KGG enriched peptides, and unmodified peptides from Human Lysate.

\subsection*{Human-1mix-TMT - ubiquitination}

\subsection*{Mouse-2mix-TMT - phosphorylation}

Here an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri). \cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, and phosphorylation in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours). Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications. Two mixtures using 11-plex were ran over the six conditions. The six conditions were split between 11 channels leading to the experimental design being unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.

\subsection*{Human label free quantification - no global profiling run}

This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 over expression (USP30 OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides. A corresponding mixed effects model was fit per modification and global protein as described previously. In contrast to the other biological experiments assessed in this paper, this experiment did not not have a separate global profiling run for the unmodified protein. In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run, however this generally leads to substantial less matches between modified and unmodified peptides, in addition to low feature counts for unmodified peptides.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

\subsection*{Existing statistical methods for experiments targeting PTMs}

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with limma were also proposed.\cite{Zhu} While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).\cite{Breitwieser:2013} Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs. Further details on existing methods can be seen in Supplementary \ref{supp-sec:ttest}.

\subsection*{Proposed approach}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Statistical modeling and parameter estimation}

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. In addition mixture may be included for features acquired via tandem mass tag (TMT) methods. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Detection of changes in PTMs}
\label{sec:adjustment_summary}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Missing value imputation}
\label{sec:missing_imp}

$MSstatsPTM$ gives the ability to impute missing feature intensities if desired. When values are imputed, it is assumed they are missing for reasons of low abundance. Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model.\cite{Tukey:1977} In order to impute a feature's missing values, the feature must be present in at least one MS Run. If the feature is not present across all MS Runs, the value will be left missing. Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers. Missing value imputation is done separately for both the PTM and global protein datasets.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Extension to TMT experiments}

The statistical modeling approaches discussed above can also be extended to Tandem Mass Tag (TMT) labeling methods. TMT experiments introduce an additional source of variation in the form of different mixtures. To account for mixture variance, first feature intensities are summarized again using Tukey's Median Polish and missing values are imputed, if desired, then a new linear mixed effects model with an added term for the mixture is included. The workflow and statisical model described generally follow the methods used for modeling proteins in $MSstatsTMT$. \cite{Huang:2020} These methods are repeated to quantify both modified and unmodified peptide abundance. Once both modified and unmodified peptides are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods described previously.

@@

\subsection*{Implementation}

The proposed methods are implemented in the open source R package $MSstatsPTM$, available on Bioconductor. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$. Conversion is done separately for the modified and global profiling runs and combined together for input into the rest of the package. If the global profiling run is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.

After using the converters, the next step is peptide/protein summarization and missing value imputation. For the modified run, the package will only summarize features, PSMs, which include the same modification at the same site. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. As mentioned in Section \ref{sec:missing_imp}, the package uses an AFT model to impute missing values in this step, although this step is optional.

The final step of the package is modeling the summarized dataset. A linear mixed effects model is automatically fit for both the summarized modified and global profiling runs. This model is automatically adjusted depending on the experimental design and acquisition method. The comparisons of interest can either be predefined or a full pairwise comparison will be tested. After modeling the modified and unmodified data, the modified model is adjusted for changes in unmodified protein abundance, using the methods described in Section \ref{sec:adjustment_summary}.

Beyond the core conversion, summary, and modeling functionality, the package also includes two plotting functions. The first creates plots for the summarization data to help with quality assurance analysis and identifying sources of variation. This function plots a quality control plot, summarizing the peptide abundance per run in the form of a boxplot, and profile plot, plotting each feature and the overall feature summarization as a line plot over each run. The second function plots the modeling results and includes a volcano plot, showing all peptides adjusted pvalues and fold changes, as well as a heatmap, which evaluates the fold change between conditions and peptides.

The package relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}, depending on the data acquisition type. The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.  All datasets in this paper were evaluated using the $MSstatsPTM$ package.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Evaluation}

The performance of of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. For the experiments where ground truth is known, we calculated summary statistics using the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). To evaluate the performance of different methods, we considered the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and total accuracy ((TP + TN) / (TP + TN + FP + FN)). For biological experiments where ground truth was not known, we evaluated significantly differential peptides with an $\alpha$ of .05. Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots.

\subsubsection*{Computer simulation 1 - label free}

In this simulation, not accounting for changes in unmodified peptide level resulted in a high false discovery rate for all methods, seen in Figure \ref{fig:fdr_all_models}. When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR. Recall and accuracy of the proposed approach and limma performed similarly, while t-test lagged behind. Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach and limma leveraged all available information, which resulted in improved power with increased number of conditions. Further results are provided in Supplementary Figures \ref{supp-fig:sim1_fpr} to \ref{supp-fig:sim1_recall}.

\subsubsection*{Computer simulation 2 - missing values and low replicates}

Changes in unmodified peptide level still needed to be accounted for to control the FDR in all methods. Once controlling for changes in unmodified peptide, it is clear that the proposed method performed stronger across all methods when compared to limma and t-test, as seen in Figure \ref{fig:sim_accuracy}. The proposed method well calibrated model accuracy and outperformed the other methods. Additionally, in the recall rate the proposed method showed the strongest performance, even when the number of biological replicates was low. Finally, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, while the other methods showed much wider distributions. Further results are provided in Supplementary Figures \ref{supp-fig:sim2_recall} to \ref{supp-fig:fc_boxplot}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{SpikeIn benchmark - KGG enriched}

@@

In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using limma with protein adjustment. Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied. However, using limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue. In this case, using limma would have resulted in missing the majority of differential modified peptides.  In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer afterward.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Human-1mix-TMT - ubiquitination}

@@

\subsubsection*{Mouse-2mix-TMT - phosphorylation}

@@

\subsubsection*{Human label free quantification - no global profiling run}

As discussed previously, there was no unmodified global profiling run performed in this experiment. Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a significant lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified protein and could be adjusted. Not having a separate global profiling run resulted in very low feature counts for the unmodified protein model.

@@

\subsubsection*{Sample size calculation}

\todo{Devon: Update this section in main and supp}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

@@

Overall, the proposed approach balances accuracy and practicality, and enables the analysis of complex experiments in high throughput. Future work is to carry out the inference and testing for not only the relative change of PTM abundance, but also the fraction of the protein that is modified at the particular site (site occupancy, or stoichiometry). We are also interested in characterizing the interplay of PTMs at multiple sites. The proposed statistical methods are implemented as an R package $MSstatsPTM$ available on Bioconductor and Github.
